Assessing primitives performance\\ on multi-stage execution

### 2018-05-18 paper.tex

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
	\item There is a big performance gap between the baseline performance and the peak performance.
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower~\cite{VMHotCold}.
\end{itemize}

@@

\begin{figure}[bth!]
		\centering
		\includegraphics[width=0.65\linewidth]{figures/VMCompilation}
		\caption{Cog VM compilation}
		\label{fig:VMCompilation}
\end{figure}

@@

\begin{figure}[tbh]
		\centering
		\includegraphics[width=0.99\linewidth]{figures/TwoStacks}
		\caption{Client and C stacks.}
		\label{fig:TwoStacks}
\end{figure}

@@

 In this section we compare the four different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). %\sd{why not ST, Plugin, VM?} %\sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM. We count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential on large string comparison:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\end{itemize}

@@

   \end{subfigure}
   \subcaption*{\textbf{\LLVMMac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. The higher the better.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline
  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.05 $\pm$ 0.0088 & 23.49 $\pm$ 0.0135 \\
  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.624 $\pm$ 0.0012 & 1.581 $\pm$ 0.0092 & 11.41 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
   \hline
  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.24 $\pm$ 0.0515 & 22.54 $\pm$ 0.0775\\
 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.690 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.15 $\pm$ 0.0321  & 12.94 $\pm$ 0.0177 \\
   \hline
  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.66 $\pm$ 0.0481 & 24.93 $\pm$ 0.0309\\
 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.64 $\pm$ 0.0600  & 24.88 $\pm$ 0.1983 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

@@

These results can be explained based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}:

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 57,8 (\GCCLinux) to 61,4\% (\LLVMMac) slower than the baseline's for empty strings. The problem is even more important for the plugin version, since the beginning of the primitive code, before the main comparison loop, is slower due to the calls to the main VM code.
As the string size goes larger, the overhead is less significant because the execution time becomes dominated by the time spent in the main comparison loop.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

\paragraph{Optimization potential on large strings comparison.}
The Plugin comes with an overhead due the use of an interface. Besides, cross-file inlining is not performed for plugins, like said in Section~\ref{sec:implem_cmp}. These are some of the factors explaining the performance gap between these versions and the \legSlang and \legRTL ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). However, the switch overhead is still noticeable in the Slang implementation when strings of 100 characters are compared: Slang is 17.2 times faster than Baseline while Slang+JIT is 18,6 times faster than Baseline. 
This last observation only applies to the \GCCLinux results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 22.54 and 24.93 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when the \LLVMMac set-up is used, with respectively 12,94 and 24,88 relative speed-up factors.

@@

This is an example of one native function in the whole VM and we do not conclude anything about the C compiler capabilites from it. VMs compiled with LLVM and GCC are around equally fast except in narrow cases like this one and LLVM is able to optimize correctly the \legSlangPlug version. LLVM likely fails at optimizing the \legSlang version likely due to the old C extensions we use in the interpreter code present in the same file to force variables into specific registers, which are better supported by GCC, historically the only compiler we supported. However, we can see that the \legRTL version has the most reliable performance since it does not depend on what optimization the C compiler is able to perform or not. This is even more relevant in less common back-ends not shown in this evaluation, such as the MIPSEL back-end we support, where C compilers are usually less clever at performing back-end specific optimizations since less engineering time has been invested into such optimizations.

One can also notive a significant gap between the Plugin and Slang implementations in \LLVMMac. This can also be explained by a difference in optimizations. As seen in Figure \ref{fig:Asm}, the comparison loop is indeed optimized in the Plugin version and is similar to the JIT implementation.

In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation. The results showed the \legRTL implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack switching, optimization for large strings, and the control over generated code.

>>>>

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
	\item There is a big performance gap between the baseline performance and the peak performance.
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
\end{itemize}

@@

\begin{figure}[bth!]
		\centering
		\includegraphics[width=0.62\linewidth]{figures/VMCompilation}
		\caption{Cog VM compilation}
		\label{fig:VMCompilation}
\end{figure}

@@

\begin{figure}[tbh]
		\centering
		\includegraphics[width=0.96\linewidth]{figures/TwoStacks}
		\caption{Client and C stacks.}
		\label{fig:TwoStacks}
\end{figure}

@@

 In this section we compare the four different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). %\sd{why not ST, Plugin, VM?} %\sk{Yes and No answers to the "recompilation needed" question, so maybe we should keep it that way}
	\item \emph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
	\item \emph{Number of lines of code:} We measured the number of lines of code of each implementation, which arguably can be used to evaluate the complexity level of each implementation.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM. We count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\end{itemize}

@@

   \end{subfigure}
   \subcaption*{\textbf{\LLVMMac}\vspace{-0.2cm}} %change vspace based on fig position
    \end{subfigure}
   \caption{String comparison benchmarks for strings of different sizes on different architectures. Average relative speed-up regarding baseline. The higher the better.}
   \label{fig:bench}
\end{figure*}
\begin{table*}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|}
  \hline
    & & 0 & 1 & 5 & 100 & 1000 \\
\hline

  \multirow{2}{*}{\legSlangPlug} &\GCCLinux& 0.413 $\pm$ 0.0005 & 0.746 $\pm$ 0.0008 & 1.804 $\pm$ 0.0005 & 12.05 $\pm$ 0.0088 & 25.36 $\pm$ 0.1742 \\
  						&\LLVMMac& 0.339 $\pm$ 0.0007 & 0.624 $\pm$ 0.0012 & 1.581 $\pm$ 0.0092 & 11.41 $\pm$ 0.0474  & 23.21 $\pm$ 0.1789 \\
   \hline
  \multirow{2}{*}{\legSlang} &\GCCLinux& 1.287 $\pm$ 0.0048 & 2.257 $\pm$ 0.0116 & 4.958 $\pm$ 0.0111 & 17.24 $\pm$ 0.0515 & 25.43 $\pm$ 0.1621\\
 					 &\LLVMMac& 0.932 $\pm$ 0.0015 & 1.690 $\pm$ 0.0030 & 3.241 $\pm$ 0.0067 & 10.15 $\pm$ 0.0321  & 12.94 $\pm$ 0.0177 \\
   \hline
  \multirow{2}{*}{\legRTL}  &\GCCLinux& 3.343 $\pm$ 0.0211 & 5.171 $\pm$ 0.0265  & 8.715 $\pm$ 0.0221 & 17.66 $\pm$ 0.0481 & 26.76 $\pm$ 0.1861\\
 				 	&\LLVMMac & 3.255 $\pm$ 0.0079 & 5.216 $\pm$ 0.0153 & 8.971 $\pm$ 0.0359 & 17.64 $\pm$ 0.06  & 24.88 $\pm$ 0.1983 \\
  \hline
\end{tabular} 
\vspace{0.2cm}
\caption{Benchmark results with standard errors. Average relative speed-up regarding baseline.}
\label{tab:detail_bench}
\end{table*}

@@

These results can be explained based on the last 3 criteria detailed in Section~\ref{sec:implem_cmp}.

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure~\ref{fig:bench} for small strings (size 1 or less): the Plugin implementation is between 58,7 (\GCCLinux) to 66,1\% (\LLVMMac) slower than the baseline's for empty strings.The problem is even more important for the \legSlangPlug version, since the beginning of its code, before the main comparison loop, is slower due to the calls to the main VM code.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, \legRTL version is always the fastest as it is not impacted by the switch overhead.

\paragraph{Optimization potential.}
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure~\ref{fig:Asm}). The stack switch overhead is still noticeable in the \legSlang implementation when strings of 100 characters are compared: in \GCCLinux, Slang is 17.2 times faster than Baseline while \legRTL is 18,6 times faster than Baseline. Due to the beginning of its code calling the VM code, the \legSlangPlug version is still around 0.68 times slower than the \legSlang and \legRTL version for a string of size 100 (Still in \GCCLinux). For strings of size 1000, the performance of all primitives is very similar, since they have almost the same comparison loop, except for the \legSlang version on \LLVMMac, as we are about to discuss in the next paragraph.

\paragraph{Control over generated code.} One can notice a rather striking performance difference between the 2 set-ups we have.In \GCCLinux, the performance between the \legSlang and \legRTL implementations is really close for long strings (1000 characters wide), with respectively 25.24 and 26.76 relative speed-up factors for a 1000-characters string. Yet performance greatly differs when the \LLVMMac set-up is used, with respectively 12,94 and 24,88 relative speed-up factors.

@@

This is an example of one native function in the whole VM and we do not conclude anything about the C compiler capabilites from it. VMs compiled with LLVM and GCC are around equally fast except in narrow cases like this one and LLVM is able to optimize correctly the \legSlangPlug version. LLVM likely fails at optimizing the \legSlang version due to the old GNU C extensions we use in the interpreter code present in the same file to force variables into specific registers, which are better supported by GCC, the only compiler we supported back in the days.

However, we can see that the \legRTL version has the most reliable performance since it does not depend on what optimization the C compiler is able to perform or not. This is even more relevant in less common back-ends not shown in this evaluation, such as the MIPSEL back-end we support. On such uncommon processors, C compilers are usually less clever at performing back-end specific optimizations since less engineering time has been invested into it. Cog's RTL is closer to the native instructions and can therefore often generate better code in this case.

\subsection*{Conclusion}In this section, we compared the performance of the different implementations against a baseline (\emph{i.e., }{pure Smalltalk code}) implementation. The results showed the \legRTL implementation was always the fastest. Otherwise, most of the performance gaps between the primitive versions could be explained by 3 criteria: stack switching, optimization for large strings, and the control over generated code.
